Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-05-24 analysis/paper_draft2.Rmd
Update paper_draft2.Rmd
edit methods and discussion

The posterior distribution does not have a closed form.  Accordingly, we used the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. To measure convergence, we used trace plots, the Gelman-Rubin convergence diagnostic and Geweke’s diagnostics [@Gelman1992;@Gelman1998;@Geweke1992].

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny].

@@

Furthermore, the bias detection rate depends on the distance between the bias parameters for the current trial and for all the past (historical) trials ($ = |b_{curr} - b_{hist}|$). As we explained above when we let $b_{hist}$ differ from $b_{curr}$ (our parameter of interest) in our simulations, these bias parameters may be incompatible ($b_{curr} \neq b_{hist}$) for at least two reasons.

@@

The simulations further show that bias detection increases with the number of an attorney's strikes per trial or the number of historical trials. This is in part because we chose a conservative initial prior for the bias parameter ($b \sim N(0,2)$) to account for _Batson_ challenger's burden of proof. This initial prior accords with @Kadane2021 [p. 51], who suggests a prior for estimating _Batson_ strike bias that is at least unimodal, symmetric, and not dependent on the data. Given little historical strike data for an attorney and a low number of strikes per trial, this initial prior will dominate, and, as a result, the model we use will only detect severe bias.

>>>>

The posterior distribution does not have a closed form.  Accordingly, we used the No U-Turn Sampler (NUTS) implemented in Rstan (v 2.21.2, Stan Dev Team, 2020). To measure convergence, we used trace plots, the Gelman-Rubin convergence diagnostic and Geweke’s diagnostics [@Gelman1992;@Gelman1998;@Geweke1992]. For each replicate, we generate an MCMC sample of 10000 iterations with a burn-in period of 1000 iterations of a single chain.

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny]. In contrast to the simulations, the application was built using Rcpp (version and citation), and the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution.

@@

Furthermore, the bias detection rate depends on the distance between the bias parameters for the current trial and for all the past (historical) trials (i.e., $ |b_{curr} - b_{hist}|$). As we explained above when we let $b_{hist}$ differ from $b_{curr}$ (our parameter of interest) in our simulations, these bias parameters may be incompatible ($b_{curr} \neq b_{hist}$) for at least two reasons.

@@

The degree of incompatability between current and historical trial data may also influcence the selection of the value of $\alpha$. For example, when there is perfect agreement between current and historical data ($ |b_{curr} - b_{hist}| = 0 $), $\alpha$ should be at its maximum (equal weight between current and historical data). This will maximize the model's ability to detect bias, even when the number of strikes is small. This effect of increased bias detection can be seen in the simulation plots above, along the positive diagonals of the individual blocks, and is strengthened by increasing the number of historical trials and number of strikes [new figure here].

The simulations further show that bias detection increases with the number of an attorney's strikes per trial or the number of historical trials, even when there is incompatability between current and historical trials. This is in part because we chose a conservative initial prior for the bias parameter ($b \sim N(0,2)$) to account for _Batson_ challenger's burden of proof. This initial prior accords with @Kadane2021 [p. 51], who suggests a prior for estimating _Batson_ strike bias that is at least unimodal, symmetric, and not dependent on the data. Given little historical strike data for an attorney and a low number of strikes per trial, this initial prior will dominate, and, as a result, the model we use will only detect severe bias. As @Kadane2021 notes, bias detection will sensitive to the choice of the standard deviation of the initial prior. Decreasing the magnitude of the this parameter (e.g., $ < 2$), will make the detection of bias more difficult, make the model more conservative, and require more evidence of bias from the likelihood (the current data).
